### 2020-04-20 README.Rmd
Finished editing Analysis, improved the presentation of some figures, added alternate models for reference, for sensitivity analysis. Next step: Results and discussion, and conclusions.

# A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

@@

>https://github.com/paezha/covid19-environmental-correlates

Context, Data, and Methods
============

@@

Explain the rationale for selecting the variables.

For example: The literature about COVID-19 suggested that population density is the one of the most important proliferate cause of these viscous, however this ill spread with different intensity at big cities of the world. Controlling for socioeconomic characteristics the objective of this paper is observe the effect of clime on COVID-19  proliferation.

Some questions are important to point out. Is not evident present a diary association between the temprature (resp. humidity) with declarate case. There is evidences (ref) that a time lag is necessary between the day that the individuo inffect and the case is diagnosticate

@@

```{r descriptive-statistics, echo=FALSE, cache=TRUE}
data.frame(Variable = c("COVID-19 Incidence",
                        "Area", 
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "Area of province in sq.km",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                    mean(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                    mean(provinces_spain$GDPpc),
                    mean(provinces_spain$Older),
                    mean(provinces_spain$Density),
                    mean(covid19_spain$Mean_Temp),
                    mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                  sd(provinces_spain$GDPpc),
                  sd(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                  sd(provinces_spain$Older),
                  sd(provinces_spain$Density),
                  sd(covid19_spain$Mean_Temp),
                  sd(covid19_spain$Humidity)),
           Source = c("Montera34",
                      "INE",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable(#"latex",
        "html",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics} Descriptive statistics") %>%
  #kable_styling(latex_options = c("striped", "scale_down")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(7, width = "5em") %>%
  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

@@

\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
-->

@@

Equation \ref{eq:sur-sim}) can be rewriten in compact form:

<!--
\begin{equation}
y = X \beta + \epsilon
\end{equation}
-->

@@

As is the case with cross-sectional data, it is possible to test the residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial]. When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation]. In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor. Spatial analytical approaches were used to understand contagion-difussion processes in the case of infectious disease in general [e.g., @Cliff1998detecting] and the 2003-2004 SARS outbreak in particular [e.g., @Meng2005understanding; @Cao2010spatio]. While we are mindful of the same caveat that the novel SARS-CoV2 may not follow the patterns of previous diseases, there is still evidence from the United States that COVID-19 displays spatial patterns that are consistent with a diffusion process [@Desjardins2020rapid]. For this reason, the spatial lag model is appropriate to model incidence of COVID-19 geographically, since it accounts for potential spatial patterns that result from a process of contagion, as explained next.

The stack expresion for the spatial lag SUR model (SUR-SLM) is as follows:

<!--
\begin{equation}
\label{eq:sur-slm}
\bf{A}y = X \beta + \epsilon \\
\epsilon =N(0,\Omega)
\end{equation}

\noindent where $A =I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.

This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces. Coefficients for the spatially lagged variable are estimated for each time period $\lambda_t$ and identifies the intensity and the sign of the impacts of neighbourhood. It is possible test the null hypotheis of identtical levels of spatial dependence ($\lambda_i=\lambda_j, \forall i,j$). The correspond Wald test is available in the `R` package `spsur`. 
-->

![\begin{equation} \label{eq:sur-slm} \bf{A}y = X \beta + \epsilon \\ \epsilon =N(0,\Omega) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbf%7BA%7Dy%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5COmega)%20%5Cend%7Bequation%7D)

\noindent where ![$A = I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3D%20I_%7BTN%7D-%5Cbf%7B%5CLambda%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5CLambda%7D%20%3D%20diag(%5Clambda_1%2C%20%5Ccdots%2C%20%5Clambda_T)%24).

This specification assumes that incidence in a province (![$y_st$](https://render.githubusercontent.com/render/math?math=%24y_st%24)) at time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average ![($Wy_{st}$)](https://render.githubusercontent.com/render/math?math=(%24Wy_%7Bst%7D%24)) of incidence in neighbouring provinces. Coefficients for the spatially lagged variable are estimated for each time period ![$\lambda_t$](https://render.githubusercontent.com/render/math?math=%24%5Clambda_t%24) and identifies the intensity and the sign of the impacts of neighbourhood. It is possible test the null hypotheis of identtical levels of spatial dependence (![$\lambda_i=\lambda_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Clambda_i%3D%5Clambda_j%2C%20%5Cforall%20i%2Cj%24)). The correspond Wald test is available in the `R` package `spsur`.

The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019). Another alternative methodologies could be use. By example, a dynamic spatial panel methodology with fixed spatial an temporal effects (e.g. Elhorst 2014, Cap. 4), but those models are more dont take account correlation between errors. Therefore, a spatial SUR approach is more reasonable for our purpose.

Analysis and Results
===================

@@

We begin with the exploratory analysis of the data.

Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of the disease, as well as its temporal progression in weekly averages. Our analysis begins on March 13, which is the first date when every province had reported at least one case of COVID-19. It can be seen that the highest incidence at this early date was in the provice of Álava, in the North of Spain. While not the most populous province, with a population of only 331,549, Álava has the highest GDP per capita of all provinces. Vitoria, its main city, is the capital of the Basque country and has been the focus of efforts to develop a "Global Basque City" [@Meijers2008strategic], along with San Sebastian and Bilbao. The other early focus of the pandemic in Spain was Madrid, which is the most populous province in the country and has the second highest GDP per capita after Álava. The early outbreaks in these two provinces can be traced throughout the progression of the pandemic over time, although by the end of the period under consideration, other provinces had matched and even surpased their incidence rates, including Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete to the south.

```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people (map omits Islas Canarias)", warning=FALSE, cache = TRUE}
# Retrieve the coordinates of Madrid
madrid_xy <-  st_centroid(provinces_spain %>% filter(province == "Madrid")) %>% st_coordinates()

week11.plot <- covid19_spain %>%
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 11) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 13 - March 15") +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 4.9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 3.4, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week12.plot <- covid19_spain %>%
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 12) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds",
                       direction = 1) +
  ggtitle("March 16 - March 22") +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 4.9, y = madrid_xy[2] + 0.6) +
  geom_segment(x = madrid_xy[1] - 3.4, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])

week13.plot <- covid19_spain %>%
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 13) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 23 - March 29") +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid",
               x = madrid_xy[1] - 4.9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 3.4, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week14.plot <- covid19_spain %>%
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 14) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 30 - April 5") +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 4.9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 3.4, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

>>>>

# A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

@@

https://github.com/paezha/covid19-environmental-correlates/tree/master/Environmental-Correlates-of-COVID19-Spain

Context and Data
============

@@

The global emergence of infectious diseases is mostly driven by environmental, ecological, and socio-economic factors [@Jones2008global]. In the case of SARS-CoV2, the ecological factors include the interaction between humans and wildlife. Once transmission of a disease begins to happen between humans, socio-economic and environmental factors become increasingly important. As noted in the introduction, the focus of the paper is on environmental variables, namely temperature, humidity, and sunshine. These variables have been implicated in the viability and ease of transmission of similar viruses. In addition to these variables, we also aim to use a set of controls, in the form of specific socio-economic and demographic characteristics of each province.

The first variable that we consider is GDP per capita. Much has been said about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p. A18)}. The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when the cause of a disease may still be unclear, but the plane has already departed [@Zhou2016accelerated]. In reference to the earlier outbreak of SARS, van Wagner [@vanWagner2008toward] chronicles how Toronto's status as a global city turned out to be a vulnerability in this respect. In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and its potential to be further ahead in the trajectory of the pandemic. Furthermore, wealthier regions also tend to concentrate more activities that produce non-traded goods, including building and construction [@Hallet2002regional]. Therefore, it is possible that wealthier regions remain relatively more active even during a lockdown. On the other hand, it is also possible that less wealthy regions have a higher proportion of workers in manual occupations that cannot telework, and therefore have more difficulties complying with shelter-in-place orders.

The percentaje of older adults (over 65) is the second variable that we consider as a control. Early evidence regarding COVID-19 suggests that the case rate mortality is higher at older ages [e.g. @Novel2020epidemiological]. However, it is not clear that a relatively large population of older adults necessarily translates into higher transmission of the infection. The tool of choice in containing the spread of the disease has been social distancing. In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility]. In other words, many older adults are, whether by preference or otherwise, already in a form of social isolation. Social distancing during the pandemic may actually reinforce that condition for them, as suggested by the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age]. Accordingly, our expectation is that provinces with higher percentages of older adults will tend to have similarly low levels of social contact, particularly since the age-structured matrix of social contact in Spain is similar to Italy [@Prem2017projecting].

The third population variable the we consider is population density, since it directly affects the contact patterns and contact rates between individuals in a population [@Hu2013scaling]. The evidence available suggests a positive relationship between the transmission of COVID-19 and population density (e.g. cumulative incidence in urban areas like NYC). For this reason, we anticipate a positive relationship between population density and the incidence of the disease.

The last control variable is the presence of mass transit systems in a province. Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility (e.g. metro or subway), namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia. Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches and simulation [e.g., @Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing. After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.

@@

```{r descriptive-statistics, echo=FALSE, cache=TRUE}
data.frame(Variable = c("COVID-19 Incidence",
                        "Area", 
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "Area of province in sq.km",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                    mean(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                    mean(provinces_spain$GDPpc),
                    mean(provinces_spain$Older),
                    mean(provinces_spain$Density),
                    mean(covid19_spain$Mean_Temp),
                    mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                  sd(provinces_spain$GDPpc),
                  sd(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                  sd(provinces_spain$Older),
                  sd(provinces_spain$Density),
                  sd(covid19_spain$Mean_Temp),
                  sd(covid19_spain$Humidity)),
           Source = c("ProvidencialData19",
                      "INE",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable(#"latex",
        "html",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics} Descriptive statistics") %>%
  #kable_styling(latex_options = c("striped", "scale_down")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(7, width = "5em") %>%
  footnote(general = c("ProvidencialData19: https://www.datoscovid.es/pages/providencialdata19", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

@@

\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t$ is a $N \times k_t$ matrix of the $k_t$ independent variables, with associated vector of coefficients $\beta_t$,; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.
-->

@@

Equation (\ref{eq:sur-sim}) can be rewriten in compact form:

<!--
\begin{equation}
\bf{Y} = \bf{X} \beta + \epsilon
\label{eq:sur-sim-block}
\end{equation}

\noindent where $\bf{Y}$ is now a vector of dimension $NT \times 1$, $\bf{X}$ is a block-diagonal matrix $NT \times K$ 
(with $K = \sum_t{k_t}$) and $\epsilon$ is an $NT \times 1$ vector. Using the Kronecker product notation ($\otimes$), the error matrix structure is expressed concisely as:

\begin{equation}
E[\epsilon \epsilon']=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'})
\end{equation}
-->

@@

\noindent where ![$\bf{Y}$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BY%7D%24) is now a vector of dimension ![$NT \times 1$](https://render.githubusercontent.com/render/math?math=%24NT%20%5Ctimes%201%24), ![$\bf{X}$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BX%7D%24) is a block-diagonal matrix ![$NT \times K$](https://render.githubusercontent.com/render/math?math=%24NT%20%5Ctimes%20K%24) 
(with ![$K = \sum_t{k_t}$](https://render.githubusercontent.com/render/math?math=%24K%20%3D%20%5Csum_t%7Bk_t%7D%24)) and ![$\epsilon$](https://render.githubusercontent.com/render/math?math=%24%5Cepsilon%24) is an ![$NT \times 1$](https://render.githubusercontent.com/render/math?math=%24NT%20%5Ctimes%201%24) vector. Using the Kronecker product notation (![$\otimes$](https://render.githubusercontent.com/render/math?math=%24%5Cotimes%24)), the error matrix structure is expressed concisely as:

![\begin{equation} E\[\epsilon \epsilon'\]=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'}) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20E%5B%5Cepsilon%20%5Cepsilon'%5D%3D%5CSigma%20%5Cotimes%20I_N%20%3B%20%5C%20%5CSigma%3D(%5Csigma_%7Btt'%7D)%20%5Cend%7Bequation%7D)

As is the case with cross-sectional data, it is possible to test the residuals of Model (\ref{eq:sur-sim-block}) for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial]. When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation]. In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor. Spatial analytical approaches were used to understand contagion-difussion processes in the case of infectious disease in general [e.g., @Cliff1998detecting] and the 2003-2004 SARS outbreak in particular [e.g., @Meng2005understanding; @Cao2010spatio]. While we are mindful of the same caveat that the novel SARS-CoV2 may not follow the patterns of previous diseases, there is still evidence from the United States that COVID-19 displays spatial patterns that are consistent with a diffusion process [@Desjardins2020rapid]. For this reason, the spatial lag model is appropriate to model incidence of COVID-19 geographically, since it accounts for potential spatial patterns that result from a process of contagion, as explained next.

The stack expresion for the SUR model with a spatially lagged dependent variable (SUR-SLM) is as follows:

<!--
\begin{equation}
\label{eq:sur-slm}
\begin{aligned}
\bf{AY} = \bf{X} \beta + \epsilon \\
\epsilon =N(0,\bf{\Sigma})
\end{aligned}
\end{equation}

\noindent where $\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$ is the spatially lagged dependent variable, and $\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$.

This specification assumes that outcome ($y_{st}$) at location $s$ and time $t$ is partially determined by the weighted average ($Wy_{st}$) of the outcome in neighboring provinces, with neighborhood defined by matrix $W$ of spatial weights. In other words, the spatially lagged dependent variable represents a process of contagion, where the disease in neighboring provinces can spillover in a spatial way. The coefficients of the spatially lagged variable are estimated for each time period $\rho_t$ and identify the intensity and the sign of the contagion effect. It is possible test the null hypothesis of identical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$). The correspond Wald test is available in the `R` package `spsur`.
-->

![\begin{equation} \label{eq:sur-slm} \begin{aligned} \bf{AY} = \bf{X} \beta + \epsilon \\ \epsilon =N(0,\bf{\Sigma}) \end{aligned} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbegin%7Baligned%7D%20%5Cbf%7BAY%7D%20%3D%20%5Cbf%7BX%7D%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5Cbf%7B%5CSigma%7D)%20%5Cend%7Baligned%7D%20%5Cend%7Bequation%7D)

\noindent where ![$\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BA%7D%20%3DI_%7BTN%7D-%5Cbf%7B%5CGamma%7D%20%5Cotimes%20W%24) is the spatially lagged dependent variable, and ![$\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7B%5CGamma%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).

This specification assumes that outcome (![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24)) at location ![$s$](https://render.githubusercontent.com/render/math?math=%24s%24) and time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average (![$Wy_{st}$](https://render.githubusercontent.com/render/math?math=%24Wy_%7Bst%7D%24)) of the outcome in neighboring provinces, with neighborhood defined by matrix ![$W$](https://render.githubusercontent.com/render/math?math=%24W%24) of spatial weights. In other words, the spatially lagged dependent variable represents a process of contagion, where the disease in neighboring provinces can spillover in a spatial way. The coefficients of the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identify the intensity and the sign of the contagion effect. It is possible test the null hypothesis of identical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)). The correspond Wald test is available in the `R` package `spsur`.

The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019).

Analysis
===================

@@

Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of COVID-19 in Spain, as well as the temporal progression of the disease in weekly averages. Our analysis begins on March 13, which is the first date when every province had reported at least one case of COVID-19. It can be seen that the highest incidence at this early date was in the provice of Álava, in the North of Spain. While not the most populous province, with a population of only 331,549, Álava has the highest GDP per capita of all provinces. Vitoria, its main city, is the capital of the Basque Country and has been the focus of efforts, along with San Sebastian and Bilbao, to develop a "Global Basque City" [@Meijers2008strategic]. The other early focus of the pandemic in Spain was Madrid, which is the most populous province in the country and has the second highest GDP per capita after Álava. The early outbreaks in these two provinces can be traced throughout the progression of the pandemic over time, although by the end of the period under consideration, other provinces had matched and even surpased their incidence rates, including Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete to the south.

```{r weekly-average-incidence-map, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:weekly-average-incidence-map}Mean weekly incidence of COVID-19 by province, in reported cases by 100,000 people", warning=FALSE, cache = TRUE}
# Retrieve the coordinates of Madrid
madrid_xy <-  st_centroid(provinces_spain %>% filter(province == "Madrid")) %>% st_coordinates()

week11.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 11) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence), color = "gray69") +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 13 - March 15") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week12.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 12) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence), color = "gray69") +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds",
                       direction = 1) +
  ggtitle("March 16 - March 22") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 9, y = madrid_xy[2] + 0.6) +
  geom_segment(x = madrid_xy[1] - 6, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])

week13.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 13) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence), color = "gray69") +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 23 - March 29") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid",
               x = madrid_xy[1] - 9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week14.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 14) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence), color = "gray69") +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 30 - April 5") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week15.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 15) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence), color = "gray69") +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("April 6 - April 11") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])
